Measuring starspot properties by forward modeling IGRINS spectra of \name

### 2016-07-15 document/ms.tex
Include or mark todos for comments from Greg Herczeg

For a single presumably coeval stellar cluster, pre main-sequence HR diagrams exhibit large ($\delta t_{\ast} \sim t_{\ast}$) apparent spreads in age \citep[e.g.][]{2011A&A...534A..83R}. The cause of the large spread in pre main-sequence HR diagrams is controversial: some take it as evidence for intrinsic age spreads, some take it as evidence for non static accretion history \citep{2009ApJ...702L..27B, 2010ARA&A..48..581S}.

Still others take the large apparent age spreads in pre main-sequence HR diagrams as evidence for the limited physics included in the pre main-sequence evolutionary model isochrones.  Most modern evolutionary models do not yet include the effects of starspots or magnetic fields, despite observations betraying their presence, \emph{e.g.} optical monitoring and Zeeman doppler imaging \citep{2008A&A...479..827G,2014MNRAS.444.3220D}.

@@

Sunspots are probably responsible for biases in stellar effective temperatures derived by different methods.  For example, the APOGEE spectrograph ($1.5-1.70 \;\mu$m at $R=22,500$) measured effective temperatures for 3493 young stars finding offsets in $\teff$ of typically 200$-$500 K and as high as 1000 K compared to previous studies \citep{2014ApJ...794..125C}.  Low resolution optical spectroscopy shows a typical spread of 200 K in the spectral type-to-effective-temperature conversion scale \citep{2014ApJ...786...97H}.  These discrepancies in $\teff$ could be attributable to measurements of spectral features steming from patches of stellar photosphere with different effective temperatures than the bulk photosphere.  Single temperature photospheres are not realistic approximations for complex young photospheres possessing large areal coverage fractions of starspots.

Stellar evolution models including the effect of starspots can make a coeval 10 Myr population exhibit apparent age spreads of 3$-$10 Myr, with derived masses biased towards lower masses \citep{2015ApJ...807..174S}.  The importance of starspots needs to be empirically evaluated to assess systematic biases of pre-main sequence stellar evolutionary models.
%...develop this point more

Modeling the periodic light curves of spotted stars can answer these questions: ``What evidence is there to support the existence of starspots?  What joint constraint can be placed upon the effective temperature contrast and areal coverage of the starspots?  What is the longitudinal asymmetry of the spots?''.  Cyclical light curve variations indicate only the longitudinally asymmetric component of the star spots \citep{2012AJ....143....4H}.  Photometric modeling of the light curve taken alone can only yield degenerate estimates in the temperature contrast and areal coverage fraction.  A star possessing longitudinally symmetric distribution of starspots, or a pole-on star with a single spot, or a non-rotating (or very slowly rotating) star would all show zero or secular amplitude of photometric variability, yet they all contain star spots.

By modeling the composite \emph{spectrum} of a spotted star, it is possible to directly measure the collective areal coverage and effective temperature contrast.  Spectroscopic modeling can still measure the starspot properties in pole-on stars, stars with longitudinally symmetric stars, and non-rotating stars.

% Previous efforts to derive starspot properties via spectroscopy:

%% Put in text here!

In principle, more advanced questions could be asked about the properties of the spots: ``How many distinct spot groups are there?  Is there evidence for both hot spots and cool spots?  What is the longitudinal distribution of the spots?  What is the (co-)latitudinal distribution of the spots?  What is the distribution of effective temperature within a spot or spot group?  What are the morphologies of the spots?''.  Solving these questions would require a complex model comparison framework, combining all available data and lines of evidence from photometric, spectroscopic, and spectropolarimetric monitoring.  We seek to answer the much easier questions about the aggregate properties of the stellar photosphere, assuming the star is well described by two distinct photospheric components.

%\subsection{Previous studies of \name}

@@

The source has little or no accretion ($\log{\dot M} < 8.1$), no detection of mid-infrared nor mm excess, and no stellar companion, so its spectrum should be devoid of complicating factors like near-IR excess veiling and accretion excess.  This source is the ideal candidate for direct measurement of two-temperature photosphere due to starspots.
% These claims need references.

@@

The details of the methodology are contained in Appendix \ref{methods-details}.  Here we overview the salient concepts needed to follow the results and discussion.

We experimented with several extensions to the modular framework for spectral inference described in \iancze \footnote{The open source codebase and its full revision history is available at https://github.com/iancze/Starfish.  The experimental fork discussed in this paper is at https://github.com/gully/Starfish}.  The \iancze technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We use the \PHOENIX~ model grid. After experimenting with the base model from \iancze, we made two principal changes to the modular framework: 1) the introduction of a starspot mixture model characterized by a second $\teff$ and areal fill factor $f$, and 2) the switch from Gibbs sampling to ensemble sampling with chunking by spectral order.  We discuss these two changes below.

Following and extending the notation of \iancze, we can write down the mixture model as:

where $\vM$ is the pixel-level forward model of a spectrum, and we assume the stellar photosphere is characterized as two photospheric components with different effective temperatures, $\teff$ and solid angular coverages $\Omega$, but sharing all the same intrinisic and extrinsic stellar parameters otherwise.  The observed filling factor of photosphere $b$ is simply the ratio:

@@

The second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \cite{2013PASP..125..306F}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations expected among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that we obtain unique stellar parameter estimates $\vT_{o}$ for each spectral order $o$, whereas the \iancze strategy had the power to provide a single set of stellar parameters $\vT$ that was based on all $N_{\rm ord}$ spectral orders.  There is no principled way to combine the $N_{\rm ord}$ sets of inferences on $\vT_{o}$, but there are ample heuristics like taking the weighted average of the $o$ sets of marginalized posterior distributions functions.

@@

\subsection{ESPaDOnS Spectroscopy}
We combined our IGRINS spectroscopy with existing ESPaDOnS panchromatic optical echelle spectra to provide 88 spectral orders between 5100 and 25000 \AA \citep{2014MNRAS.444.3220D}.  There are 12 spectra observed over the 13 days between JD 2456665.72 and 2456678.90.

@@

There is a $\Delta \teff \sim 800$ K difference in the effective temperature derived at different wavelength intervals for the same stellar spectrum.  To put this level of effective temperature in context, authors using 4200 K and 3300 K values would conclude evolutionary model derived masses of XX or YY as shown in Figure XX. 
\todo{Put in the numbers and HR diagram figure here.}

@@

\emph{Suitability of pre-computed \PHOENIX model spectra, model ranges, $\A \neq 0$, and magnetic fields-} We have assumed that the \PHOENIX synthetic spectra are a good representation to our observed IGRINS spectrum of \name, a young star likely to possess a large magnetic field.  Furthermore, cool sunspots are associated with heightened magnetic fields on the sun.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high Lande-g factors (cite XX, fix the g formatting).  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants (cite XX Saar and collaborators).  Our tacit assumption has been that these effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures and 2) there exist many orders and lines that are not affected at all.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$ due to Zeeman broadening.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.
% Note that these two effects would differently affect the spectra

@@

\begin{tabular}{ r| p{5cm} | p{5cm} | }
\multicolumn{1}{r}{}
 &  \multicolumn{1}{c}{Whole spectrum (Gibbs Sampling)}
 & \multicolumn{1}{c}{Order-by-order (emcee)} \\
\cline{2-3}
Single component photosphere & \citet{2015ApJ...812..128C}: Poor model for spotted stars & Sec. \ref{sec:approach12}- plausible evidence for starspots if $\teff$ varies with order \\
\cline{2-3}
Two component photosphere & \ref{sec:approach21}- Statistically principled inference on starspot properties given the whole spectrum.  But strongly correlated dimensions impose challenge on MCMC tuning. & \ref{sec:approach22}- Multiple independent posteriors of starspot properties.  But combining inferences is heuristic.  \\
\cline{2-3}
\end{tabular}

>>>>

\todo{Overhaul the introduction: 1. Move discussion about age spreads later,  2. }

For a single presumably coeval stellar cluster, pre main-sequence HR diagrams exhibit large ($\delta t_{\ast} \sim t_{\ast}$) apparent spreads in age \citep[e.g.][]{2011A&A...534A..83R}. The cause of the large spread in pre main-sequence HR diagrams is controversial: some take it as evidence for intrinsic age spreads, some take it as evidence for non static accretion history \citep{2009ApJ...702L..27B, 2010ARA&A..48..581S}.  Still others take the large apparent age spreads in pre main-sequence HR diagrams as evidence for the limited physics included in the pre main-sequence evolutionary model isochrones.  Most modern evolutionary models do not yet include the effects of starspots or magnetic fields, despite observations revealing their presence, \emph{e.g.} optical monitoring and Zeeman doppler imaging \citep{2008A&A...479..827G,2014MNRAS.444.3220D}.

@@

Sunspots are probably responsible for biases in stellar effective temperatures derived by different methods.  For example, the APOGEE spectrograph ($1.5-1.70 \;\mu$m at $R=22,500$) measured effective temperatures for 3493 young stars finding offsets in $\teff$ of typically 200$-$500 K ($\sim$0.02$-$0.05 dex) and as high as 1000 K ($\sim$0.1 dex) compared to previous studies \citep{2014ApJ...794..125C}.  Low resolution optical spectroscopy scales between that of \citet{2003ApJ...593.1093L} and those of \citet{2013ApJS..208....9P} and \citet{2014ApJ...786...97H}, showing about 200 K in the spectral type-to-effective-temperature conversion scale.  These discrepancies in $\teff$ could be attributable to measurements of spectral features steming from patches of stellar photosphere with different effective temperatures than the bulk photosphere.  Single temperature photospheres are not realistic approximations for complex young photospheres possessing large areal coverage fractions of starspots.

\comm{From GH regarding the next paragraph: Starspots (or magnetic activity, or whatever) doesn't cause a spread in ages/evolution.  A \emph{range} in starspots may cause a \emph{range} in evolution, but that paper requires a sample.  
I think we can refer to this sort of thing, but we shouldn't focus on it.}
Stellar evolution models including the effect of starspots can make a coeval 10 Myr population exhibit apparent age spreads of 3$-$10 Myr, with derived masses biased towards lower masses \citep{2015ApJ...807..174S}.  The importance of starspots needs to be empirically evaluated to assess systematic biases of pre-main sequence stellar evolutionary models.

\todo{Discuss limitations and capabilities of photometry alone: presence or absence of starspots, joint constraint on areal coverage and temperature, longitudinal asymmetry.}
Cyclical light curve variations indicate only the longitudinally asymmetric component of the star spots \citep{2012AJ....143....4H}.  Photometric modeling of the light curve taken alone can only yield degenerate estimates in the temperature contrast and areal coverage fraction.  A star possessing longitudinally symmetric distribution of starspots, or a pole-on star with a single spot, or a non-rotating (or very slowly rotating) star would all show zero or secular amplitude of photometric variability, yet they all contain star spots.

By modeling the composite spectrum of a spotted star, it is possible to directly measure the collective areal coverage and effective temperature contrast.  Spectroscopic modeling can still measure the starspot properties in pole-on stars, stars with longitudinally symmetric stars, and non-rotating stars.

\todo{Cite previous efforts to derive starspot properties via spectroscopy}

\todo{More advanced approaches to understanding stellar surface properties.  How many distinct spot groups are there?  Is there evidence for both hot spots and cool spots?  What is the longitudinal distribution of the spots?  What is the (co-)latitudinal distribution of the spots?  What is the distribution of effective temperature within a spot or spot group?  What are the morphologies of the spots?}
In principle, more advanced questions could be asked about the properties of the spots....

Solving these questions would require a complex model comparison framework, combining all available data and lines of evidence from photometric, spectroscopic, and spectropolarimetric monitoring.  We seek to answer the much easier questions about the aggregate properties of the stellar photosphere, assuming the star is well described by two distinct photospheric components.

\todo{Previous studies of \name}

@@

The source has no detected accretion, no detection of mid-infrared nor mm excess, and no detected stellar companion, so its spectrum should be devoid of complicating factors like near-IR excess veiling and accretion excess.  This source is the ideal candidate for direct measurement of two-temperature photosphere due to starspots.
\todo{These claims need references.}

@@

\iancze developed a modular framework to infer stellar properties from high resolution spectra.  In this section, we describe how we extended this framework to include spots in a 2-temperature fit to spectra.  Further details of this methodology are described in Appendix \ref{methods-details}.

We experimented with several extensions to the modular framework for spectral inference described in \iancze \footnote{The open source codebase and its full revision history is available at https://github.com/iancze/Starfish.  The experimental fork discussed in this paper is at https://github.com/gully/Starfish}.  The \iancze technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We use the \PHOENIX~ grid of pre-computed synthetic stellar spectra that span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest\citep{2013A&A...553A...6H}.  \todo{Put in a hyperlink?} After experimenting with the base model from \iancze, we made two principal changes to the modular framework: 1) the introduction of a starspot mixture model characterized by a second $\teff$ and areal fill factor $f$, and 2) the switch from Gibbs sampling to ensemble sampling with chunking by spectral order.  We discuss these two changes below.

We assume the stellar photosphere is characterized as two photospheric components with different effective temperatures, $\teffa$ and $\teffb$, and scalar solid angular coverages $\Omega_a$ and $\Omega_b$, but sharing all the same intrinisic and extrinsic stellar parameters otherwise.  Following and extending the notation of \iancze, we can write down the mixture model as:

\begin{eqnarray} \label{eqn:mix_M}
\vM_{\mathrm{mix}} = \Omega_a \vM(\teffa)  + \Omega_b \vM(\teffb)
\end{eqnarray}

The observed filling factor of photosphere $b$ is simply the ratio:

@@

The second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \cite{2013PASP..125..306F}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations expected among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that we obtain unique stellar parameter estimates $\vT_{o}$ for each spectral order $o$, whereas the \iancze strategy had the power to provide a single set of stellar parameters $\vT$ that was based on all $N_{\rm ord}$ spectral orders.  We combine the $N_{\rm ord}$ sets of inferences on $\vT_{o}$ with heuristics like taking the weighted average of point estimates of parameter distributions of the most reliable subsets of orders.  We assume that these lossy heuristics offer a regression to the mean with a large number of spectral orders.

@@

\subsection{ESPaDOnS Spectroscopy}
We retrieved twelve existing ESPaDOnS spectra from \citet{2014MNRAS.444.3220D}.
\todo{Say some more about the spectra- what wavelength range used, chunking.}

@@

The effective temperature derived at different wavelength intervals exhibit a $\Delta \teff \sim 800$ K spread.  To put this level of effective temperature in context, authors using 4200 K and 3300 K values would conclude evolutionary model derived masses of XX or YY as shown in Figure XX. 
\todo{Put in the numbers and HR diagram figure here.}

@@

\emph{Suitability of pre-computed \PHOENIX model spectra, model ranges, $\A \neq 0$, and magnetic fields-} We have assumed that the \PHOENIX synthetic spectra are a good representation to our observed IGRINS spectrum of \name, a young star likely to possess a large magnetic field.  Cool sunspots are associated with heightened magnetic fields on the sun.  Magnetic fields will have two effects on the emergent spectrum- line broadening attributable to the Zeeman effect, and pressure broadening attributable to seeing deeper into deeper layers in the photosphere compared to the surrounding hotter photosphere.  The former effect will differentially impact lines with high Lande-g factors (cite XX, fix the g formatting).  The latter effect could masquerade as an apparent shift in surface gravity, as seen in mixture models of spotted stars with empirical templates composed of dwarfs and giants (cite XX Saar and collaborators).  Our tacit assumption has been that these effects are 1) relatively small compared to the gross appearance and disappearance of spectral lines at disparate temperatures and 2) there exist many orders and lines that are not affected at all.  We see tentative evidence for physics outside of the pre-computed model grids in the poor fits to the observed spectra in $K-$band.  Strong metal lines demonstrate large departures in their best fit $\Z$ due to Zeeman broadening.  No pre-computed model grids including a range of surface magnetic fields are available at the wavelength range and spectral sampling needed to approach these questions within this spectral inference framework.
% Note that these two effects would differently affect the spectra

@@

\todo{Simplify this discussion to what matters here.}
